---
title: "Exploring Bikeshare data"
author: "Khadidja"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Project Overview
Exploring data related to bike share systems for three major cities in the United States—Chicago, New York City, and Washington.
We will be answering interesting questions about it by computing descriptive statistics and making visualizations!

# Q01 - Popular times of travel | What is the most common month?


```{r}
#Importing data
chi<-read.csv('chicago.csv')
ny<-read.csv('new-york-city.csv')
wash<-read.csv('washington.csv')

#Loading some libraries we are in need to
library(lubridate)  #Manipulating the dates
library(modeest)  #Calculating mode
library(ggplot2)  #Creating visualizations
library(dplyr)  #Manipulating data frames

#Creating new column only for months in each city data set
chi$month <- month(chi$Start.Time) 
ny$month <- month(ny$Start.Time)
wash$month <- month(wash$Start.Time)

#Creating function 'Common_month', input: short and complete city name, returns: the most common month + data visualization
common_month <- function(cit,pseudo){
  result <- (paste("The most common month of travel in",pseudo,"is the month number:",mfv(cit$month)))  #mfv() finds the most frequent value
  plot <- ggplot(aes(x = month), data = cit) +
    geom_histogram(binwidth = 0.5, color = I('black'), fill = I('darkcyan')) +
    scale_x_continuous(breaks = seq(1,6,1)) +
    ggtitle(paste('The monthly travels histogram for',pseudo,'city')) +
    xlab ('Months of travel') +
    ylab ('Count of travels')
  
  list1 <- list(result,plot)
  return(list1)
}

```

```{r}
#Using the function common_month, inserting the city names as arguments
common_month(chi,"Chicago")
```



The histogram shows clearly that 'June' holds the highest number of travels, about 100000 travels, followed by 'May' with about 66000 travels, while the smallest number about 21900 is for 'January'
```{r}
common_month(ny,"New York")
```

For New York city 'June' is also the most common with more than 75000 travels followed by 'May', while 'January' holds the least number of travels about 32000.
```{r}
common_month(wash,"Washington")
```


'June' is the most common for Washington as well, with more than 68000 travels, followed by 'April' with more than 62000 travels, and the smallest number about 30000 is assigned for 'January'.

<font size="4"> Summary statistics </font>

```{r}
#New data frames holding the months and the count of travels
C <- aggregate(chi$month, by = list(chi$month), FUN = length) #Chicago
N <- aggregate(ny$month, by = list(ny$month), FUN = length) #New York
W <- aggregate(wash$month, by = list(wash$month), FUN = length) #Washington

#Summary Stats on the months
summary(C$x)  #Chicago
summary(N$x)  #New York
summary(W$x)  #Washington
```

The actual values resulting from the statistical calculations confirm exactly what we had appearing on the data visualizations above.
The max and the min values confirm the ranges we had from the charts.

## Q02 - Popular stations and trip | What is the most common start station?


```{r}
#Creating new function, input: short and complete name, returns: the most common Start Station + data visualization + the summary stats related to the top 5 stations
common_station <- function(cit,pseudo){
  result1 <- paste("The most common start station for",pseudo,"city is",mfv(cit$Start.Station))
  
  df <- aggregate(cit$Start.Station, by = list(cit$Start.Station), FUN = length)#Creating new data frame storing the station with the related count of travels
  df <- tail(df[order(df$x),],5) #Pulling the top 5 stations after ordering the data frame
  names(df)[1] <- "Start_station" #Renaming the columns
  names(df)[2] <- "Count"
  
  result2 <- summary(df$Count)
  
  result3 <- ggplot(aes(x = Start_station, y = Count),data = df) +
    geom_bar(stat = 'identity', width = 0.5, color = I('black'), fill = I('darkseagreen1')) +
    ggtitle(paste('Top 05 Start stations in',pseudo,'city')) +
    ylab ('Count of travels') +
    coord_flip()
  
  list2 <- list(result1,result3,result2)
  return(list2)
}
```

```{r}
#Using the function common_station, inserting the city names as arguments
common_station(chi,"Chicago")
```

The previous bar chart compares the counts of travels of the top 5 Start stations in Chicago.
'Streeter Dr & Grand Ave' is the most common, followed by 'Clinton St & Washington Blvd'. The counts range from about 3400 to 6900.

```{r}
common_station(ny, "New York")
```

The bar chart for New York shows that 'Pershing Square North' is the most common followed by 'E 17 St & Broadway'.
The number of travels ranges from about 1900 to 3000.
```{r}
common_station(wash, "Washington")
```

'Columbus Circle / Union Station' is the most common station followed by 'Lincoln Memorial' and the last one is '15th & P St NW'.
The values range between about 5600 and 3500.

The results of summary stats integrated in the function confirm exactly the intervals of values taken from the plots, for all the cities. 

## Q03 - Trip duration | What is the total travel time for each user type in different cities?

<font size="5"> Starting with 'Chicago' </font> 

```{r}
#Applying the sum function using the by command, getting each user type with the total travel time in minutes 
by(chi$Trip.Duration/60,chi$User.Type, sum) #divide Trip.Duration by 60 to get the time in minutes
```

```{r}
#As long as the total travel time for 'Dependent' type is insignificant, we will be plotting only for 'Customer' and 'Subscriber'
newdf <- chi[-which(chi$User.Type == 'Dependent'),]  #Creating new data frame newdf, 'Dependent' omitted 
qplot(x = User.Type, y = Trip.Duration/60, data = newdf,
      geom = 'boxplot', color = I("blue4"), fill = I('deepskyblue3')) +
  coord_cartesian(ylim = c(0,40)) +  #Adjusting and omitting the outliers
  ggtitle("Travel time per user type in Chicago city") +
  labs(x = "User type",y = "Trip duration in minutes")
```

The median line is much higher for Customers, which indicates that Users of type 'Customer' registered longer travel time than 'Subscribers'.
50% of Customers spent about 23 minutes while the Subscribers spent about 9 minutes.  

<font size="4"> Summary stats</font> 

```{r}
by(newdf$Trip.Duration/60,newdf$User.Type, summary)
```
By looking at the actual values obtained we can confirm our results on the graph.
The first quartile at 25% for Customers is 15.45 and looks right on the plot.
The 75% mark for Subscribers is 14.983 and it is exactly what had showing on the boxplot.


<font size="5">'New York' city</font> 
```{r}
#Finding the total travel time for each user type
by(ny$Trip.Duration/60,ny$User.Type, sum)
```
 
```{r}
#Notice that we have a third type but has no name, we will keep only "Customer" and "Subscriber" for the upcoming explorations
newdf1 = ny[-which(ny$User.Type == ''),] #new data frame, unnamed type removed

qplot(x = User.Type, y = Trip.Duration/60, data = newdf1,
      geom = 'boxplot', color = I("cadetblue4"), fill = I('cadetblue3')) +
  coord_cartesian(ylim = c(0,40)) +  #Adjusting and omitting the outliers
  ggtitle("Travel time per user type in New York city") +
  labs(x = "User type",y = "Trip duration in minutes")
```

From the plot we observe that the Interquartile Range for "Customers" lays between 14 and 29 mins, and the median is at about 22 mins.
For "Subscribers" the IQR is between 6 and 16 mins, and its median falls at about 9 mins.
The difference between them is considerable that the time for Subscribers is almost half of that for Customers. 

<font size="4"> Summary stats</font> 
```{r}
by(newdf1$Trip.Duration/60,newdf1$User.Type, summary)
```
Again the exact calculation above proves the results obtained from the graph, IQR for Customer between 14.23 and 29.15 mins, IQR for Subscriber between 5.817 and 15.683 mins 

<font size="5">'Washington' city</font> 
```{r}
#Finding the total travel time for each user type
by(wash$Trip.Duration/60,wash$User.Type, sum)
```
```{r}
qplot(x = User.Type, y = Trip.Duration/60, data = wash,
      geom = 'boxplot', color = I("darkslategray3"), fill = I('darkslategray2')) +
  coord_cartesian(ylim = c(5,45)) +  #Adjusting and omitting the outliers
  ggtitle("Travel time per user type in Washington city") +
  labs(x = "User type",y = "Trip duration in minutes")
```

The difference between Customer and Subscriber in time is always very large, that the third quartile  for Subscribers is slightly under the first quartile for Customers at about  15 mins.

<font size="4"> Summary stats</font> 
```{r}
by(wash$Trip.Duration/60,wash$User.Type, summary)
```

And finally to check on the exactness of the data appearing on the graph, the 25% mark for Customers is 15.334 and the 75% mark for Subscribers is 15.045 mins, which is very true!

<center> Thanks for your Time :3 </center>
